Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-11 ms.qmd
some more writing

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Dewitt1998; @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020; @Dewitt1998]. The implications of changes in variance could therefore have wide-reaching implications for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

@@

@@

Understanding acclimation capacity and how the opportunity for selection will change across populations and species is important for predicting the impacts of climate change [@Chevin2017; @Chevin2010]. Here, we show that the opportunity for selection is expected to decrease for wild ectotherms, but particularly so for terrestrial ectotherms (~28%). Decreases in variance in physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the opportunity for selection for marine and terrestrial ectotherms (between 4-13% reduction in total variance) under climate change assuming a fraction of this variation is heritable and/or it weakens selection.

### **Acclimation capacity and opportunity for selection is limited in response to climate change**
One explanation for why terrestrial ectotherms show minimal acclimation capacity may be related to the fact that terrestrial ectotherms, were, on average acclimated for significantly less time than ectotherms from aquatic habitats. However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have higher rates of acclimation than fishes; See @Einum2023-lz]. Faster rates of acclimation would therefore result in the opposite patterns to those we observed -- in other words, terrestrial species would be more likely to exhibit lower $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ when controlling for acclimation time.

@@

### ***Ecological consequences of reduced variation in physiological Rates***
- need analysis of traits for lnCVR.

>>>>

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Dewitt1998; @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Bolnick2011 and @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020; @Dewitt1998]. Such changes in physiological trait variation can may also have important ecological consequences by promotong population productivity and stability [@Agashe2009], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011]. The implications of changes in variance could therefore have wide-reaching consequences for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

@@

Traits did not differ in whether they exhibited a reduction in variation in physiological rates at high temperatures. Each life-history stage also exhibited the same pattern of variance change in each of the habitats (@fig-cvrlh).

```{r, fig-cvrlh}
#| label: fig-cvrlh
#| fig-cap: "Predicted mean $lnCVR_{Q_{10}}$ for adult and juevenile life-history stages for A) Marine, B) Freshwater and C) Terrestrial ectotherms."
#| echo: false
#| warning: false

p1 <- orchard_plot(model2.4, mod = "type", at = list(c_accl_time = 0, habitat = "m"), group = "record_num", xlab = TeX("$lnCVR_{Q_{10}}$"), condition.lab = "Life Stage", by = "life.hist",  g = FALSE, k = FALSE, legend.pos = "top.left", angle = 45) + labs(title = "Marine") + ylim(c(-2, 2.2))

p2 <- orchard_plot(model2.4, mod = "type", at = list(c_accl_time = 0, habitat = "f"), group = "record_num", xlab = TeX("$lnCVR_{Q_{10}}$"), condition.lab = "Life Stage", by = "life.hist",  g = FALSE, k = FALSE)+ labs(title = "Freshwater") + theme(legend.position = "none", axis.text.y = element_blank())

p3 <- orchard_plot(model2.4, mod = "type", at = list(c_accl_time = 0, habitat = "t"), group = "record_num", xlab = TeX("$lnCVR_{Q_{10}}$"), condition.lab = "Life Stage", by = "life.hist",  g = FALSE, k = FALSE)+ labs(title = "Terrestrial")+ theme(legend.position = "none", axis.text.y = element_blank())

fig4 <- (p1 | p2 | p3) + plot_annotation(tag_levels = "A", tag_suffix = ")")
ggsave(fig4, filename = "./output/figures/fig4_lh.png", width = 14, height = 4.5)

fig4 <- magick::image_read("./output/figures/fig4_lh.png")
fig4
```

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the impacts ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here we show that the opportunity for selection is expected to decrease for ectotherms, but particularly so for terrestrial ectotherms (~28%). Decreases in variance in physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the variability in physiologcial rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections.

### **Acclimation capacity and opportunity for selection is limited in response to climate change**
One explanation for why terrestrial ectotherms show minimal acclimation capacity may be related to the fact that terrestrial ectotherms, were, on average acclimated for significantly less time than ectotherms from aquatic habitats. However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have higher rates of acclimation than fishes; See @Einum2023-lz]. Faster rates of acclimation would therefore result in the opposite patterns to those we observed -- in other words, terrestrial species would be more likely to exhibit lower $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ when controlling for acclimation time. 
assuming a fraction of this variation is heritable and/or it weakens selection

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability, particularly if oscillations are predictable over time [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016]. Empirical evidence properly testing these predictions, however, is scant. @Leung2020 recently showed that greater plasticty in morphological traits did evolve in experimental algae populations, however, work in seed beetles. Our results suggest that such theretical explanations may not be general but rather specific to the population, context and trait in question.

### ***Ecological consequences of reduced variation in physiological Rates***
The ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remains unclear. The dearth of empirical data demonstrating changes in variance paired with a lack of theoretical models 
- buckley 2014, - Lawton 1991, - Nisbet
